---
title: "Multinomial Classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multinomial Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo = F}
body {
  text-align: justify;
}

p {
  font-weight: 'normal';
  font-size: '16px;
  text-align: justify;
}

p.note {
  font-weight: 'normal';
  font-size: '12px;
  font-family: 'Serif';
  text-align: justify;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<p>This example presents the application of the PieGlyph package to show the predicted probabilities of different classes in a multinomial classification problem. This example could be useful for clustering problems which gives probabilities, e.g. `mclust`.</p>


### Load libraries
```{r setup, warning = F, message = FALSE}
library(ranger)
library(ggplot2)
library(PieGlyph)
library(dplyr)
```

### Load data

<p>We are using `iris` dataset which gives the measurements of the sepal length, sepal width, petal length and petal width for 50 flowers from each of <i>Iris setosa, Iris versicolor, </i> and <i>Iris virginica </i></p>
```{r data,warning = F, message = FALSE}
head(iris)
```

### Classification model

<p> We use the random forest algorithm for classifying the samples into the three species according to the four measurements described above.</p>

```{r model, warning = F}
rf <- ranger(Species ~ Petal.Length + Petal.Width + 
                        Sepal.Length + Sepal.Width, 
             data=iris, probability=TRUE)
```

<p>We get the predicted probabilities of each sample belonging to a particular species.</p>

```{r predictions, warning = F}
preds <- as.data.frame(predict(rf, iris)$predictions)
head(preds)
```

<p>Combine the predicted probabilities with the original data for plotting</p>
```{r plot_data, warning = F}
plot_data <- cbind(iris, preds)
head(plot_data)
```

### Create plot

```{r prob-plot, warning = F, fig.align='center', fig.width=7}
ggplot(data=plot_data,
       aes(x=Sepal.Length, y=Sepal.Width))+
  # Pies-charts showing predicted probabilities of the different species
  geom_pie_glyph(slices = names(preds), colour = 'black')+
  # Colours for sectors of the pie-chart
  scale_fill_manual(values = c('#56B4E9', '#D55E00','#CC79A7'))+
  # Labels for axes and legend
  labs(y = 'Sepal Width', x = 'Sepal Length', fill = 'Prob (Species)')+
  theme_minimal()
```

